---
title: "Food Desert Analysis"
author: ""
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=FALSE}
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 

```


```{r setup}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(readr)
library(ggthemes)
library(caret)
library(randomForest)
library(pROC)
library(car)
library(broom)
library(reshape2)
library(plotly)
library(patchwork) 
library(tidyr)
library(corrplot)
```
# Introduction
Background

Food deserts—areas with limited access to affordable and nutritious food—are a critical public health concern in the United States. According to USDA definitions, food deserts are typically low-income census tracts where a substantial number or share of residents have low access to supermarkets or large grocery stores. Our analysis examines the geographic distribution of food deserts, their socioeconomic correlates, potential health impacts, and predictive modeling approaches to identify high-risk areas.

Dataset Overview

We utilized data from the USDA Food Access Research Atlas (2019), supplemented with socioeconomic, health, and food environment datasets. The combined dataset includes:

    72,531 census tracts across the United States

    200+ variables covering food access, demographics, socioeconomic status, and health outcomes

    Key indicators of food desert status, poverty rates, vehicle access, and health statistics

Key Variables:
Variable Name	Description
LILATracts_1And10 - 	Food desert indicator (1=mile urban/10-mile rural threshold)
PovertyRate	 - Percentage of population below poverty line
TractSNAP - 	Percentage of households receiving SNAP benefits
PCT_OBESE_ADULTS17-	Percentage of obese adults
PCT_DIABETES_ADULTS13 -	Percentage of adults with diabetes
Urban - Urban/rural classification


```{r chunk1}
# Load dataset
raw_df <- read_csv("/Users/MANOJ/Desktop/PROJECT SUMMARY/FoodAccessResearchAtlasData2019.csv")

# Print dataset columns
str((raw_df))
```
```{r}

df_fixed <- raw_df %>%
  mutate(across(where(is.character), ~ na_if(., "NULL")))

# Step 2: Now calculate missing value percentages
missing_pct <- colMeans(is.na(df_fixed))

# Step 3: Drop columns with more than 10% missing
df_clean <- df_fixed[, missing_pct <= 0.10]

str(df_clean)
```

```{r}
df_first3 <- df_clean %>% select(1:3)

# Step 2: Work on the rest of the columns
df_rest <- df_clean %>% select(-1:-3)

# Step 3: Identify binary columns (only two unique values: 0 and 1)
binary_cols <- sapply(df_rest, function(x) {
  x_unique <- unique(na.omit(x))
  length(x_unique) == 2 && all(x_unique %in% c(0, 1))
})

# Step 4: Convert binary columns to factor, others to numeric
df_rest_clean <- df_rest %>%
  mutate(across(names(binary_cols)[binary_cols], as.factor)) %>%
  mutate(across(names(binary_cols)[!binary_cols], as.numeric))

# Step 5: Combine everything back
df <- bind_cols(df_first3, df_rest_clean)

# Check the structure
str(df)


```

**SMART Q1**

Where are food deserts most geographically concentrated across the U.S., and how do these concentrations differ between urban and rural census tracts?
-----------------------------------------------------------------------------------------------------
According to the USDA definitions, a "food desert" is typically a low-income tract 
that also has low access to supermarkets based on established distance criteria.
Here, we use the 'LILATracts_1And10' column (which applies a 1-mile threshold for urban
and a 10-mile threshold for rural areas) as an indicator.
We assume that a value of 1 in 'LILATracts_1And10' indicates that the tract qualifies as a food desert.
# Methodology
 Used USDA-defined food desert indicator (LILATracts_1And10)
 Analyzed state-level patterns using maps
 Compared urban vs. rural distributions


```{r chunk2}
df$FoodDesert <- as.numeric(as.character(df$LILATracts_1And10))

state_urban_counts <- df %>%
  group_by(State, Urban) %>%
  summarise(FoodDesert = sum(FoodDesert, na.rm = TRUE), .groups = 'drop')

food_desert_pivot <- state_urban_counts %>%
  pivot_wider(names_from = Urban, values_from = FoodDesert, values_fill = 0) %>%
  rename(`Rural Food Desert Count` = `0`, `Urban Food Desert Count` = `1`) %>%
  mutate(`Total Food Desert Count` = `Rural Food Desert Count` + `Urban Food Desert Count`) %>%
  arrange(desc(`Total Food Desert Count`))

head(food_desert_pivot, 10)
```



```{r}
state_totals <- df %>%
  group_by(State) %>%
  summarise(
    TotalFoodDeserts = sum(FoodDesert, na.rm = TRUE)
  ) %>%
  arrange(desc(TotalFoodDeserts))  # Sort descending for a nice plot


state_totals_top10 <- head(state_totals, 10)

ggplot(state_totals_top10, aes(x = reorder(State, TotalFoodDeserts), y = TotalFoodDeserts)) +
  geom_bar(stat = "identity", fill = "coral") +
  coord_flip() +
  labs(
    title = "Top 10 States by Number of Food Deserts",
    x = "State",
    y = "Number of Food Deserts"
  ) +
  theme_minimal(base_size = 14)
```





```{r chunk3}
# Summarize food desert percentage by state
state_summary <- df %>%
  group_by(State) %>%
  summarise(
    TotalFoodDeserts = sum(FoodDesert, na.rm = TRUE),
    TotalTracts = n(),
    FoodDesertPercentage = 100 * TotalFoodDeserts / TotalTracts
  )

# Mapping state names to abbreviations
state_abbrev <- data.frame(
  State = state.name,
  Abbreviation = state.abb,
  stringsAsFactors = FALSE
)

state_summary <- left_join(state_summary, state_abbrev, by = "State")

# Add centroids for US states
state_centroids <- data.frame(
  Abbreviation = state.abb,
  latitude = state.center$y,
  longitude = state.center$x
)

state_summary <- left_join(state_summary, state_centroids, by = "Abbreviation")

# 1. Choropleth layer
fig <- plot_ly()

fig <- fig %>% add_trace(
  type = 'choropleth',
  locations = state_summary$Abbreviation,
  locationmode = 'USA-states',
  z = state_summary$FoodDesertPercentage,
  colorscale = 'Reds',
  colorbar = list(title = "Food Desert %"),
  text = paste(state_summary$State, "<br>",
               "Food Desert %: ", round(state_summary$FoodDesertPercentage, 1), "%"),
  hoverinfo = "text"
)

# 2. Add text labels properly at lat/lon
fig <- fig %>% add_trace(
  type = 'scattergeo',
  lat = state_summary$latitude,
  lon = state_summary$longitude,
  text = paste0(round(state_summary$FoodDesertPercentage), "%"),
  mode = 'text',
  textfont = list(color = "black", size = 9),
  showlegend = FALSE
)

# 3. Layout adjustments
fig <- fig %>% layout(
  title = "Percentage of Food Desert Tracts by State (2019)",
  geo = list(
    scope = 'usa',
    projection = list(type = 'albers usa'),
    showlakes = FALSE,
    lakecolor = 'white'
  )
)

# Show the plot
fig


```

```{r}
state_summary <- df %>%
  group_by(State, Urban) %>%
  summarise(
    TotalFoodDeserts = sum(FoodDesert, na.rm = TRUE),
    TotalTracts = n(),
    FoodDesertPercentage = 100 * TotalFoodDeserts / TotalTracts,
    .groups = 'drop'
  )

state_abbrev <- data.frame(State = state.name, Abbreviation = state.abb, stringsAsFactors = FALSE)
state_centroids <- data.frame(Abbreviation = state.abb, latitude = state.center$y, longitude = state.center$x)

state_summary <- state_summary %>%
  left_join(state_abbrev, by = "State") %>%
  left_join(state_centroids, by = "Abbreviation")

urban_summary <- filter(state_summary, Urban == 1)

# Urban Map
fig_urban <- plot_ly()

# Urban Choropleth
fig_urban <- fig_urban %>% add_trace(
  type = 'choropleth',
  locations = urban_summary$Abbreviation,
  locationmode = 'USA-states',
  z = urban_summary$FoodDesertPercentage,
  colorscale = 'Blues',
  colorbar = list(title = "Urban Food Desert %"),
  zmin = 0,
  zmax = max(state_summary$FoodDesertPercentage, na.rm = TRUE),
  text = paste(urban_summary$State, "<br>",
               "Urban Desert %: ", round(urban_summary$FoodDesertPercentage, 1), "%"),
  hoverinfo = "text"
)

# Urban text labels
fig_urban <- fig_urban %>% add_trace(
  type = 'scattergeo',
  lat = urban_summary$latitude,
  lon = urban_summary$longitude,
  text = paste0(round(urban_summary$FoodDesertPercentage), "%"),
  mode = 'text',
  textfont = list(color = "black", size = 9),
  showlegend = FALSE
)

# Urban layout
fig_urban <- fig_urban %>% layout(
  title = "Urban Food Desert % by US State",
  geo = list(
    scope = 'usa',
    projection = list(type = 'albers usa'),
    showlakes = TRUE,
    lakecolor = 'white'
  )
)

fig_urban


rural_summary <- filter(state_summary, Urban == 0)

# Rural Map
fig_rural <- plot_ly()

# Rural Choropleth
fig_rural <- fig_rural %>% add_trace(
  type = 'choropleth',
  locations = rural_summary$Abbreviation,
  locationmode = 'USA-states',
  z = rural_summary$FoodDesertPercentage,
  colorscale = 'Greens',
  colorbar = list(title = "Rural Food Desert %"),
  zmin = 0,
  zmax = max(state_summary$FoodDesertPercentage, na.rm = TRUE),
  text = paste(rural_summary$State, "<br>",
               "Rural Desert %: ", round(rural_summary$FoodDesertPercentage, 1), "%"),
  hoverinfo = "text"
)

# Rural text labels
fig_rural <- fig_rural %>% add_trace(
  type = 'scattergeo',
  lat = rural_summary$latitude,
  lon = rural_summary$longitude,
  text = paste0(round(rural_summary$FoodDesertPercentage), "%"),
  mode = 'text',
  textfont = list(color = "black", size = 9),
  showlegend = FALSE
)

# Rural layout
fig_rural <- fig_rural %>% layout(
  title = "Rural Food Desert % by US State",
  geo = list(
    scope = 'usa',
    projection = list(type = 'albers usa'),
    showlakes = TRUE,
    lakecolor = 'white'
  )
)

fig_rural

```





```{r chunk4}
state_summary <- df %>%
  group_by(State, Urban) %>%
  summarise(
    TotalFoodDeserts = sum(FoodDesert, na.rm = TRUE),
    TotalTracts = n(),
    FoodDesertPercentage = 100 * TotalFoodDeserts / TotalTracts,
    .groups = 'drop'
  )

# Separate Urban and Rural
urban_summary <- state_summary %>%
  filter(Urban == 1) %>%
  arrange(desc(FoodDesertPercentage))

rural_summary <- state_summary %>%
  filter(Urban == 0) %>%
  arrange(desc(FoodDesertPercentage))

# 2. Create Urban Plot
urban_plot <- ggplot(urban_summary, aes(x = reorder(State, FoodDesertPercentage), y = FoodDesertPercentage)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Urban Food Desert % by State",
    x = "State",
    y = "Urban Desert %"
  ) +
  theme_minimal(base_size = 12)

# 3. Create Rural Plot
rural_plot <- ggplot(rural_summary, aes(x = reorder(State, FoodDesertPercentage), y = FoodDesertPercentage)) +
  geom_bar(stat = "identity", fill = "seagreen") +
  coord_flip() +
  labs(
    title = "Rural Food Desert % by State",
    x = "State",
    y = "Rural Desert %"
  ) +
  theme_minimal(base_size = 12)

# 4. Combine side by side
urban_plot + rural_plot + plot_layout(ncol = 2)
```

By far Texas has the most food desert tracts, but it also has a large number of total tracts.
Mississippi has the highest percentage of food desert tracts, with an astonishing 32% of its tracts being food deserts.
Mississppi also has the highest percentage of food desert tracts in urban areas which contribute to 42% of its total tracts.
Arizona and alaska share the highest percentage of food desert tracts in rural areas, with 29%.


---------------------------------------------------------------------------------------------


**SMART Q2**
What demographics and socioeconomic factors are linked to food deserts?
# Methodology
Feature selection using univariate screening, LASSO, and VIF analysis
Logistic regression modeling
Correlation analysis of top predictors

```{r chunk5}

foodatlas <- read_csv("/Users/MANOJ/Desktop/PROJECT SUMMARY/FoodAccessResearchAtlasData2019.csv",col_types = cols(CensusTract = col_character()))
socioeconomic <- read_csv("/Users/MANOJ/Desktop/PROJECT SUMMARY/FE_socioeconomic.csv")
insecurity <- read_csv("/Users/MANOJ/Desktop/PROJECT SUMMARY/FE_insecurity.csv")
health <- read_csv("/Users/MANOJ/Desktop/PROJECT SUMMARY/FE_health.csv")
stores <- read_csv("/Users/MANOJ/Desktop/PROJECT SUMMARY/FE_stores.csv")
restaurants <- read_csv( "/Users/MANOJ/Desktop/PROJECT SUMMARY/FE_restaurants.csv")
taxes <- read_csv("/Users/MANOJ/Desktop/PROJECT SUMMARY/FE_taxes.csv")
local <- read_csv("/Users/MANOJ/Desktop/PROJECT SUMMARY/FE_local.csv")
access <- read_csv("/Users/MANOJ/Desktop/PROJECT SUMMARY/FE_access.csv")
state_data <- read_csv("/Users/MANOJ/Desktop/PROJECT SUMMARY/FE_supplemental_data_state.csv")
county_data <- read_csv("/Users/MANOJ/Desktop/PROJECT SUMMARY/FE_supplemental_data_county.csv")

# 2. Preprocessing

# Replace "NULL" strings with NA across all datasets
clean_nulls <- function(df) {
  df %>% mutate(across(where(is.character), ~ na_if(., "NULL")))
}

foodatlas <- clean_nulls(foodatlas)
socioeconomic <- clean_nulls(socioeconomic)
insecurity <- clean_nulls(insecurity)
health <- clean_nulls(health)
stores <- clean_nulls(stores)
restaurants <- clean_nulls(restaurants)
taxes <- clean_nulls(taxes)
local <- clean_nulls(local)
access <- clean_nulls(access)
state_data <- clean_nulls(state_data)
county_data <- clean_nulls(county_data)

# 3. Merge datasets

# Ensure all FIPS are properly padded
foodatlas <- foodatlas %>%
  mutate(CensusTract = str_pad(CensusTract, 11, pad = "0"),
         CountyFIPS = substr(CensusTract, 1, 5))

socioeconomic <- socioeconomic %>% mutate(CountyFIPS = str_pad(FIPS, 5, pad = "0"))
insecurity <- insecurity %>% mutate(CountyFIPS = str_pad(FIPS, 5, pad = "0"))
health <- health %>% mutate(CountyFIPS = str_pad(FIPS, 5, pad = "0"))
stores <- stores %>% mutate(CountyFIPS = str_pad(FIPS, 5, pad = "0"))
restaurants <- restaurants %>% mutate(CountyFIPS = str_pad(FIPS, 5, pad = "0"))
taxes <- taxes %>% mutate(CountyFIPS = str_pad(FIPS, 5, pad = "0"))
local <- local %>% mutate(CountyFIPS = str_pad(FIPS, 5, pad = "0"))
access <- access %>% mutate(CountyFIPS = str_pad(FIPS, 5, pad = "0"))
county_data <- county_data %>% mutate(CountyFIPS = str_pad(FIPS, 5, pad = "0"))

# Merge everything into foodatlas
merged_df <- foodatlas %>%
  left_join(socioeconomic, by = "CountyFIPS") %>%
  left_join(insecurity, by = "CountyFIPS") %>%
  left_join(health, by = "CountyFIPS") %>%
  left_join(stores, by = "CountyFIPS") %>%
  left_join(restaurants, by = "CountyFIPS") %>%
  left_join(taxes, by = "CountyFIPS") %>%
  left_join(local, by = "CountyFIPS") %>%
  left_join(access, by = "CountyFIPS") %>%
  left_join(county_data, by = "CountyFIPS")


merged_df$FoodDesert <- as.numeric(as.character(merged_df$LILATracts_1And10))

# 3. Select numeric predictors
predictor_vars <- merged_df %>%
  select(-FoodDesert) %>%
  select(where(is.numeric))    
  

# 4. Drop columns with >10% missing
missing_pct <- colMeans(is.na(predictor_vars))
predictor_vars <- predictor_vars[, missing_pct <= 0.10]

# 5. Impute remaining NA with medians
predictor_vars <- predictor_vars %>%
  mutate(across(everything(), ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))

is_binary <- function(x) {
  unique_vals <- unique(na.omit(x))
  length(unique_vals) == 2 && all(sort(unique_vals) %in% c(0, 1))
}

binary_cols <- names(predictor_vars)[sapply(predictor_vars, is_binary)]
continuous_cols <- setdiff(names(predictor_vars), binary_cols)

# 7. Convert binary variables to factor
predictor_vars <- predictor_vars %>%
  mutate(across(all_of(binary_cols), as.factor))

# 8. Scale continuous variables
predictor_vars <- predictor_vars %>%
  mutate(across(all_of(continuous_cols), scale))

# 6. Prepare modeling data
model_data <- bind_cols(
  FoodDesert = merged_df$FoodDesert,
  predictor_vars
) %>% drop_na()

set.seed(97)
model_data_sampled <- model_data %>% sample_frac(0.3)

str(model_data_sampled)

```





```{r}


selected_vars <- c(
  "LowIncomeTracts",
  "LILATracts_Vehicle",
  "lalowihalfshare",
  "MedianFamilyIncome",
  "lasnap10share",
  "PovertyRate",
  "LA1and10",
  "FoodDesert"
)

selected_data <- merged_df %>%
  select(all_of(selected_vars)) %>%
  mutate(FoodDesert = as.factor(FoodDesert))

selected_data_numeric <- selected_data %>%
  mutate(across(everything(), ~ as.numeric(as.character(.))))


# Calculate correlation matrix
cor_matrix <- cor(selected_data_numeric, use = "complete.obs", method = "spearman")  # Ignore missing values safely


corrplot(cor_matrix, method = "color", type = "upper", 
         addCoef.col = "black", tl.col = "black", tl.srt = 45)
```

```{r}
plots <- lapply(selected_vars[1:7], function(var) {
  ggplot(selected_data_numeric, aes(x = FoodDesert, y = .data[[var]])) +
    geom_boxplot(fill = "skyblue") +
    labs(title = var, x = "Food Desert", y = var) +
    theme_minimal(base_size = 12)
})

# Combine all plots
wrap_plots(plots, ncol = 4) 
```

**SMART Q3**
How are food deserts related to health issues like obesity and diabetes?
# Methodology
Comparative analysis of health outcomes by food desert status
Correlation analysis using Spearman method
Boxplot visualization of distributions


```{r chunk6}
health_vars <- c(
  "PCT_OBESE_ADULTS17",   # % Obese Adults
  "PCT_DIABETES_ADULTS13", # % Adults with Diabetes
  "FoodDesert"
)

# 2. Subset and prepare data
health_data <- merged_df %>%
  select(all_of(health_vars)) %>%
  mutate(FoodDesert = as.factor(FoodDesert))  # Treat FoodDesert as factor for plotting

# 3. Boxplots for each health indicator
for (var in health_vars[1:2]) {  # Loop over obesity and diabetes
  p <- ggplot(health_data, aes(x = FoodDesert, y = .data[[var]])) +
    geom_boxplot(fill = "tomato") +
    labs(
      title = paste("Boxplot of", var, "by Food Desert Status"),
      x = "Food Desert (0 = No, 1 = Yes)",
      y = var
    ) +
    theme_minimal(base_size = 14)
  
  print(p)
}
```


```{r}
health_data_numeric <- health_data %>%
  mutate(across(everything(), ~ as.numeric(as.character(.))))

# 4. Calculate the correlation matrix
cor_matrix_health <- cor(health_data_numeric, use = "complete.obs", method = "spearman")

corrplot(cor_matrix_health, method = "color", type = "upper", 
         addCoef.col = "black", tl.col = "black", tl.srt = 45)



```




```{r chunk7}
selected_vars <- c(
  "LowIncomeTracts",
  "LILATracts_Vehicle",
  "lalowihalfshare",
  "MedianFamilyIncome",
  "lasnap10share",
  "PovertyRate",
  "LA1and10"
)

# 2. Prepare the modeling data
model_data <- merged_df %>%
  select(all_of(c(selected_vars, "FoodDesert"))) %>%
  drop_na()  # Drop rows with NA values

# 3. Build logistic regression model
final_formula <- as.formula(paste("FoodDesert ~", paste(selected_vars, collapse = " + ")))

library(speedglm)
logit_model <- speedglm(final_formula, data = model_data, family = binomial())

# 4. Summarize model
summary(logit_model)

# 5. Tidy output for better view
tidy_logit <- tidy(logit_model) %>%
  mutate(
    odds_ratio = exp(estimate),
    p.value = p.value
  ) %>%
  arrange(p.value)

# 6. View top results
print(tidy_logit)




```
```{r chunk8}
set.seed(75)
train_idx <- createDataPartition(model_data$FoodDesert, p = 0.7, list = FALSE)
train_data <- model_data[train_idx,]
test_data <- model_data[-train_idx,]

rf_model <- randomForest(FoodDesert ~ ., data = train_data, ntree = 100)
rf_probs <- predict(rf_model, newdata = test_data, type = "prob")[,2]

roc_obj <- roc(test_data$FoodDesert, rf_probs)
auc(roc_obj)

rf_preds <- ifelse(rf_probs > 0.5, 1, 0)
confusionMatrix(as.factor(rf_preds), as.factor(test_data$FoodDesert))

# Variable importance
varImpPlot(rf_model)
```

#Recommendations
#Policy Implications
Targeted Interventions: Focus resources on high-risk areas identified by models
Transportation Solutions: Address vehicle access limitations in rural food deserts
SNAP Optimization: Enhance program accessibility in high-poverty regions

#Future Research

Longitudinal study design to examine causal relationships
Incorporate more granular health outcome data
Explore alternative modeling approaches (e.g., XGBoost, neural networks)
Include additional geographic and economic indicators

#Conclusion

Our analysis reveals that food deserts are most prevalent in Southern states and are strongly associated with poverty, limited transportation access, and SNAP participation. While no direct health impacts were identified, the developed predictive models effectively identify high-risk areas using socioeconomic indicators. These findings provide valuable insights for policymakers addressing food insecurity and can guide targeted interventions to improve food access in vulnerable communities.
References

#USDA Food Access Research Atlas (2019)
U.S. Census Bureau American Community Survey data - https://www.ers.usda.gov/data-products/food-access-research-atlas/download-the-data
